Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-02-21 ms.tex
mopping up appendix B

Figure~\ref{fig:stag} shows our results for the stagnation radius $r_{\rm s}$, expressed in ratio to sphere of influence radius $r_{\rm soi}$, with different colors showing different values of $v_{w}$.  Cusp and core galaxies are marked with square and triangles, respectively.  Shown for comparison are our analytic results (eq.~[\ref{eq:rs2main}]) with solid and dashed lines for cusp and core galaxies, respectively.  Our analytic estimates provide accurate fits to the numerical results for $v_{\rm w} \gtrsim \sigma$, but they break down for low $v_{\rm w}$ as the stagnation radius diverges to large radii.  For $v_{w} \gg \sigma$ the ratio $\rs/\rsoi$ does not depend sensitively on break radius $r_{b}$, but the latter has a greater influence when $v_{w} \lesssim \sigma$.  The three vertically-aligned green squares located at the same value of $v_{w}/\sigma$ close to where the solid analytic line diverges correspond to models with the same black hole mass ($\Mbh = 10^7 \Msun$) and heating rate ($v_w = 300$ km s$^{-1}$) but different values of the break radius: $r_{b} = $ 50, 100, and 400 pc (from top to bottom).  
%% AG: Describe conservation checks to additionally validate the results.

@@

Energy and mass input to the CNM by stellar winds is the sum of contributions from main sequence and post-main sequence stars.  At early times following star formation, energy input is dominated by core collapse supernovae and by the fast outflows of Wolf-Rayet stars (e.g., \citealt{VossDiehl+:2009a}).  At later times energy input is dominated by main sequence winds (e.g., \citealt{NaimanSoares-Furtado+:2013a}).  Mass input is also dominated by massive stars for very young stellar populations, but for most stellar ages the slow AGB winds of evolved low mass stars dominates the mass budget.

In Appendix \ref{app:windheat} we calculate the wind heating rate from stellar winds, $v_{\rm w}^{\star}$, and the mass loss parameter, $\eta$ (eq.~[\ref{eq:q}]), as a function of age, $\tau_{\star}$, of a stellar population which is formed impulsively (Figs.~\ref{fig:vwimp}, \ref{fig:etaImp}).  At the earliest times ($\tau_{\star} \lesssim 10^{7}$ yr), the wind heating rate exceeds 1000 km s$^{-1}$, while much later ($\tau_{\star} \sim t_{\rm h}$) stellar wind heating dominated by main sequence winds is much lower, $v^{\star}_{\rm w} \sim 50-100 $ km s$^{-1}$.  We show in $\S\ref{sec:combined}$ that for the case of quasi-continueous star formation representative of the average star formation history of loss mass galaxies, the stellar wind heating rate produced by stellar winds can also be significant, $v_{\rm w}^{\star} \gtrsim 1000$ km s$^{-1}$.  Stellar wind heating thus contributes a potentially important source of both energy and mass for the CNM.

@@

Figure~\ref{fig:vwSourcesImp} shows the contributions of various
heating sources as a function of time $\tau_{\star}$ after a burst of
star formation for a black hole masses of $\Mbh = 10^{6} \Msun$ (top
panel) and $\Mbh = 10^{8} \Msun$ (bottom panel). The stellar wind
heating is calculated as described in appendix~\ref{app:windheat}. The
and Ia heating ($v_{w}^{\rm Ia}$) and compton heating ($v_{\rm
  w}^{\bullet}$) are calculated using equations~\eqref{eq:vIa}
and~\eqref{eq:vC} respectively. Both heating sources depend on the
value of the mass return parameter, $\eta$. We have plotted this for a
burst of star formation Figure~\ref{fig:etaImp}.  The Ia heating rate
also depends be a convolution of the delay time distribution
(eq. [\ref{eq:DTD}]) and the star formation history. This convolution
reduces to the delay time distribution itself for a burst of star
formation. To account for the non-steady nature of the Ia heating we
multiply the Ia heating by $e^{-\rs/\rIa}$, where $\rIa$ is the radius
where the time between subsequent supernovae is equal to the local
dynamical time. The heating contribiution from
Ia supernovae will be suppressed if the time between successive
supernovae is much greater than the dynamical time at $\rs$. Note that
both $v_{\rm w}^{\bullet}$ and $v_{\rm w}^{\rm Ia}$ depend on the
total heating rate (all of the heating sources added in quadrature),
and so we have to solve a system of impllicit eqautions for these.
The MSP contribution is negligible for both cases and is not plotted.

>>>>

Figure~\ref{fig:stag} shows our results for the stagnation radius $r_{\rm s}$, expressed in ratio to sphere of influence radius $r_{\rm soi}$, with different colors showing different values of $v_{w}$.  Cusp and core galaxies are marked with square and triangles, respectively.  Shown for comparison are our analytic results (eq.~[\ref{eq:rs2main}]) with solid and dashed lines for cusp and core galaxies, respectively, calculated assuming $n = 1$ and $n= 0.5$, respectively.  Our analytic estimates are found to provide accurate fits to the numerical results for $v_{\rm w} \gtrsim \sigma$, but the assumptions leading to these estimates become invalid for low $v_{\rm w}$ as the stagnation radius diverges to large radii.  For $v_{w} \gg \sigma$ the ratio $\rs/\rsoi$ does not depend sensitively on break radius $r_{b}$, but the latter has a greater influence when $v_{w} \lesssim \sigma$.  The three green squares which are vertically aligned at the same value of $v_{w}/\sigma$ (close to where the analytic estimate diverges) correspond to models with the same black hole mass ($\Mbh = 10^7 \Msun$) and heating rate ($v_w = 300$ km s$^{-1}$) but different values of the break radius: $r_{b} = $ 50, 100, and 400 pc (from top to bottom).  As $r_{\rm b} \rightarrow r_{s}$, the effects of the break radius is to the push the stagnation radius inwards, decreasing the SMBH accretion rate.    
%% AG: Describe conservation checks to additionally validate the results.

@@

Energy and mass input to the CNM by stellar winds is the sum of contributions from main sequence and post-main sequence populations.  At early times following star formation, energy input is dominated by core collapse supernovae and by the fast outflows of Wolf-Rayet stars (e.g., \citealt{VossDiehl+:2009a}).  At later times energy input is dominated by main sequence winds (e.g., \citealt{NaimanSoares-Furtado+:2013a}).  Mass input is also dominated by massive stars for very young stellar populations, but for most stellar ages the slow AGB winds of evolved low-mass stars dominates the mass budget.

In Appendix \ref{app:windheat} we calculate the wind heating rate from stellar winds, $v_{\rm w}^{\star}$, and the mass loss parameter, $\eta$ (eq.~[\ref{eq:q}]), as a function of age, $\tau_{\star}$, of a stellar population which is formed impulsively (Fig.~\ref{fig:vwimp}).  At the earliest times ($\tau_{\star} \lesssim 10^{7}$ yr), the wind heating rate exceeds 1000 km s$^{-1}$, while much later ($\tau_{\star} \sim t_{\rm h}$) stellar wind heating dominated by main sequence winds is much lower, $v^{\star}_{\rm w} \sim 50-100 $ km s$^{-1}$.  We show in $\S\ref{sec:combined}$ that for the case of quasi-continueous star formation representative of the average star formation history of loss mass galaxies, the stellar wind heating rate produced by stellar winds can also be significant, $v_{\rm w}^{\star} \gtrsim 1000$ km s$^{-1}$.  Stellar wind heating thus contributes a potentially important source of both energy and mass for the CNM.

@@

Figure~\ref{fig:vwSourcesImp} shows the contributions of various
heating sources as a function of time $\tau_{\star}$ after a burst of
star formation for a black hole masses of $\Mbh = 10^{6} \Msun$ (top
panel) and $\Mbh = 10^{8} \Msun$ (bottom panel). The stellar wind
heating is calculated as described in appendix~\ref{app:windheat}. The
and Ia heating ($v_{w}^{\rm Ia}$) and compton heating ($v_{\rm
  w}^{\bullet}$) are calculated using equations~\eqref{eq:vIa}
and~\eqref{eq:vC} respectively. Both heating sources depend on the
value of the mass return parameter, $\eta$. We have plotted this for a
burst of star formation in Figure~\ref{fig:vwImp}.  The Ia heating rate
also depends be a convolution of the delay time distribution
(eq. [\ref{eq:DTD}]) and the star formation history. This convolution
reduces to the delay time distribution itself for a burst of star
formation. To account for the non-steady nature of the Ia heating we
multiply the Ia heating by $e^{-\rs/\rIa}$, where $\rIa$ is the radius
where the time between subsequent supernovae is equal to the local
dynamical time. The heating contribiution from
Ia supernovae will be suppressed if the time between successive
supernovae is much greater than the dynamical time at $\rs$. Note that
both $v_{\rm w}^{\bullet}$ and $v_{\rm w}^{\rm Ia}$ depend on the
total heating rate (all of the heating sources added in quadrature),
and so we have to solve a system of impllicit eqautions for these.
The MSP contribution is negligible for both cases and is not plotted.

### 2015-02-21 windheat.tex
mopping up appendix B

For a stellar population that forms impulsively, the rate of mass and energy injection per unit stellar mass at age $t$ are given, respectively, by 
\begin{align} 
  \dot{\bar{m}}(t) &= \frac{\Delta M(t) \mu|_{M_{\rm TO}(t)}
    \left|\dot{M}_{\rm TO}(t)\right| + f_{\rm MS} \int_{m_0}^{m_{\rm
        T}(t)}
    \dot{m}(\Mstar, t) \mu|_{\Mstar} {\rm d}\Mstar }{\bar{m}_*}\\
  \dot{\bar{e}}(t) &=\dot{e}_{\rm TO}(t)+ f_{\rm MS} \int_{m_0}^{m_{\rm T}(t)}
  \frac{\vw^2(\Mstar, t) \dot{m}(\Mstar, t) \mu|_{\Mstar} {\rm d}\Mstar}{\bar{m}_*},
  \label{eq:edotImp}
\end{align} 
where the first and second terms in each expression corresponds to
contributions from main sequence (MS) stars and post-main sequence
(PMS) stars, respectively.  Here $ f_{\rm MS} < 1$ is the efficiency
with which main sequence winds thermalize their energy with the rest
of the injected gas (we hereafter assume $f_{\rm MS} = 1$) and $\mu$
is the IMF, which we take to be of the Salpeter form $\mu(\Mstar)\sim
M^{-2.35}$, truncated at $m_0 = 0.1 \Msun$ on the low mass end and at
$100 \Msun$ on the high mass end, where $\bar{m}_*$ = 0.35 $\Msun$ is
the corresponding mean stellar mass.

The quantity of mass lost in PMS winds $\Delta M(t)$ is estimated from the expression given by \citet{CiottiOstriker:2007a} (their eq.~[10]),
\begin{align}
\Delta M=
\begin{cases}
0.945 M_{\rm TO}-0.503 & M_{\rm TO} < 9 \Msun\\
 M_{\rm TO}-1.4 \Msun &  M_{\rm TO} \ge 9 \Msun,
\end{cases}
\end{align}
where $M_{\rm TO}$ is the turn-off mass, which at time $t$ is calculated as
\begin{align}
\log(M_{\rm TO})/M_{\odot} =0.24 + 0.068 x^2-0.34 x+4.76 e^{-4.58 x},
\end{align}
where $x=\log(t/10^9 {\rm years})$.  This functional fit is designed
to reproduce the results of \citet{MaederMeynet:1987a} (their Table 9)
for massive stars while asymptoting to the formula provided by
\citet{CiottiOstriker:2007a} (their eq.~[9]) for intermediate and late
times ($t \gsim 10^8$ years). {\bf AG: Mathematical aside--this
  expression divereges for $t \gg t_{\rm hubble}$}

@@

The effective wind velocity in the limit of an impulsive star formation may then be written
as 
\begin{align}
\bar{v}_w(t)=2 \dot{\bar{e}}(t)/\dot{\bar{m}}(t)
\label{eq:vwImp}
\end{align}
while {\bf AG: eta expression is now fixed--note that
  $\dot{\bar{m}}(t)$ is the mass injection rate per stellar mass, so
  that eta is dimensionless as it should be }
\begin{align}
\eta = \dot{\bar{m}}(t) \th
\label{eq:vwImp}
\end{align}
We plot $\bar{v}_w(t)$ and $\eta(t)$ in Figures~\ref{fig:vwImp} and~\ref{fig:etaImp}.

\begin{figure}
\includegraphics[width=\columnwidth]{vwImp.pdf}
\caption{\label{fig:vwImp} The effective heating rate, $\vw^{\star}$, from stellar winds from
  a stellar population of age $\tau_{\star}$.}
\end{figure}

\begin{figure}
\includegraphics[width=\columnwidth]{etaImp.pdf}
\caption{\label{fig:etaImp} The mass loss parameter, $\eta$ (eq.~[\ref{eq:q}]), resulting from stellar winds from
  a stellar population of age $\tau_{\star}$  {\bf BDM: combine with above figure into one}.}
\end{figure}

>>>>

For a stellar population that forms impulsively, the rate of mass and energy injection per unit stellar mass at age $t$ are given, respectively, by 
\begin{align} 
  \dot{\bar{m}}(t) &= \frac{\Delta M(t) \mu|_{M_{\rm TO}(t)}
    \left|\dot{M}_{\rm TO}(t)\right| + f_{\rm MS} \int_{m_0}^{m_{\rm
        T}(t)}
    \dot{m}(\Mstar, t) \mu|_{\Mstar} {\rm d}\Mstar }{\bar{m}_*}\\
  \dot{\bar{e}}(t) &=\dot{e}_{\rm TO}(t)+ f_{\rm MS} \int_{m_0}^{m_{\rm T}(t)}
  \frac{\vw^2(\Mstar, t) \dot{m}(\Mstar, t) \mu|_{\Mstar} {\rm d}\Mstar}{\bar{m}_*},
  \label{eq:edotImp}
\end{align} 
where the first and second terms in each expression corresponds to
contributions from main sequence (MS) stars and post-main sequence
(PMS) stars, respectively.  Here $ f_{\rm MS} < 1$ is the efficiency
with which main sequence winds thermalize their energy with the rest
of the injected gas (we hereafter assume $f_{\rm MS} = 1$) and $\mu$
is the IMF, which we take to be of the Salpeter form $\mu|_{\Mstar} \propto
M_{\star}^{-2.35}$, truncated at $m_0 = 0.1 \Msun$ on the low mass end and at
$100 \Msun$ on the high mass end, where $\bar{m}_*$ = 0.35 $\Msun$ is
the corresponding mean stellar mass.

The quantity of mass lost in PMS winds $\Delta M(t)$ is estimated from the expression given by \citet{CiottiOstriker:2007a} (their eq.~[10]),
\begin{align}
\Delta M=
\begin{cases}
0.945 M_{\rm TO}-0.503 & M_{\rm TO} < 9 \Msun\\
 M_{\rm TO}-1.4 \Msun &  M_{\rm TO} \ge 9 \Msun,
\end{cases}
\end{align}
where $M_{\rm TO}$ is the turn-off mass, which at time $t < t_{\rm h}$ is calculated as
\begin{align}
\log(M_{\rm TO})/M_{\odot} =0.24 + 0.068 x^2-0.34 x+4.76 e^{-4.58 x},
\end{align}
where $x=\log(t/10^9 {\rm years})$.  This functional fit is designed
to reproduce the results of \citet{MaederMeynet:1987a} (their Table 9)
for massive stars while asymptoting to the formula provided by
\citet{CiottiOstriker:2007a} (their eq.~[9]) for intermediate and late
times ($t \gsim 10^8$ years).

@@

The effective wind velocity in the limit of an impulsive star formation may then be written
as 
\begin{align}
\bar{v}_w(t)=2 \dot{\bar{e}}(t)/\dot{\bar{m}}(t)
\label{eq:vwImp}
\end{align}
while 
\begin{align}
\eta = \dot{\bar{m}}(t) \th
\label{eq:vwImp}
\end{align}
 Figur~\ref{fig:vwImp} shows the values of $\bar{v}_w(t)$ and $\eta(t)$ as a function of stellar age, $\tau_{\star}$.

\begin{figure}
\includegraphics[width=\columnwidth]{vwImp.pdf}
\includegraphics[width=\columnwidth]{etaImp.pdf}
\caption{\label{fig:vwImp} Effective heating rate, $v_w^{\star}$, and mass loss parameter, $\eta$ (eq.~[\ref{eq:q}]), resulting from stellar winds from a stellar population of age $\tau_{\star}$.}
\end{figure}
